Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-23 optim.qmd
add interpreation to glm

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = index, group = index_name)) +
  geom_point(size = 0.5) +
  facet_nested_wrap(~index_name + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value") 
```

@tbl-mod-data combines all the variables calculated above (the JS
optimiser success rate, jellyfish hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, since none of the 50
simulations identified the sine-wave shape. A generalised linear model
with a binomial family and a logit link function is used to fit the data
and @tbl-mod-output presents the model outputs. The model suggests that
the success rate of JSO is positively associated with the two jellyfish
hyper-parameters, as well as with smoothness and squintability. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate. The variable `squintability` and `dimension`
are significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success.

@@

>>>>

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = index, group = index_name)) +
  geom_line() + 
  facet_nested_wrap(~index_name + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value") 
```

@@

@tbl-mod-data combines all the variables calculated above (the JSO success rate, JSO hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the JSO hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, because upon visual inspection of the final projection, none of the 50
simulations identified the sine-wave shape. A generalised linear model
with a binomial family and a logit link function is used to fit the data
and @tbl-mod-output presents the model outputs. The model suggests that
the JSO success rate is positively associated with the two JSO
hyper-parameters, as well as with the optimisation properties -- smoothness and squintability. Specifically, using 10 more jellyfish and 10 more tries will increase the odd ratio of success by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and `r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%, respectively. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate by `r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and `r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension`
are significant, suggesting their importance relative to JSO
hyper-parameters in the optimisation success. Users can increase the 
number of jellyfish and the number of tries to improve the success rate, until JSO exceeds a 30-second runtime.

### 2024-06-23 optim.tex
add interpreation to glm

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

\begin{equation}\protect\hypertarget{eq-optimization}{}{
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
}\label{eq-optimization}\end{equation}

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, titlerule=0mm, bottomrule=.15mm, toprule=.15mm, leftrule=.75mm, bottomtitle=1mm, breakable, toptitle=1mm, arc=.35mm, colback=white, rightrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{equation}\protect\hypertarget{eq-squintability}{}{
\varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)
}\label{eq-squintability}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-logistic}{}{
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
}\label{eq-logistic}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-parametric}{}{
f(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
}\label{eq-parametric}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-squintability-parametric}{}{
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}
}\label{eq-squintability-parametric}\end{equation}

@@

\hypertarget{sec-app-1}{%
\subsection{Performance of JSO in pipe-finding
problems}\label{sec-app-1}}

@@

\hypertarget{sec-app-2}{%
\subsection{Factors affecting JSO success rate: optimisation properties
and jellyfish hyper-parameters}\label{sec-app-2}}

@@

\end{figure}

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JS optimiser success rate, jellyfish hyper-parameters, and
optimisation properties) into one table. Three preprocessing steps are
applied to the data: 1) scaling the jellyfish hyper-parameters by a
factor of 10 for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 30
seconds, and 3) encoding the success rate for the stringy case as 0,
since none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and Table~\ref{tbl-mod-output} presents
the model outputs. The model suggests that the success rate of JSO is
positively associated with the two jellyfish hyper-parameters, as well
as with smoothness and squintability. However, being flagged with long
runtime and an increase of data dimension will reduce the success rate.
The variable \texttt{squintability} and \texttt{dimension} are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success.

\hypertarget{tbl-smoothness-squintability}{}
\begin{table}
\caption{\label{tbl-smoothness-squintability}Parameters estimated from the Gaussian process (including variance,
range, smoothness, and nugget) and scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for the pipe-finding and sine-wave
finding problems. The column ``smooth'' and ``squint'' represent the
smoothness and squintability measures. }\tabularnewline

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 3.053\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 2.960\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 2.948\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 3.117\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 1.257\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.772\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 1.322\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 1.369\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 2.947\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 2.756\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 3.118\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 2.957\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

@@

\hypertarget{tbl-mod-data}{}
\begin{longtable}[t]{lrrrrrrl}
\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed for modelling. }\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule
\end{longtable}
\endgroup{}

@@

\hypertarget{tbl-mod-output}{}
\begin{longtable}[t]{|>{}lrrr>{}r|}
\caption{\label{tbl-mod-output}Model estimates of proportion of jellyfish success on optimisation
properties and jellyfish hyper-parameters from the generalised linear
model with a binomial family and a logit link function. The variable
smoothness, squintability, number of jellyfish and maximum number of
tries are positively associated with JSO success rate while data
dimension and being flagged as long runtime are negatively associated
with the success rate. The variable squintability and dimension are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success. }\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 2.060 & 0.684 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule
\end{longtable}
\endgroup{}

\hypertarget{sec-conclusion}{%
\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}}

>>>>

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}

@@

\begin{equation}\phantomsection\label{eq-optimization}{
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
}\end{equation}

@@

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacityback=0, bottomrule=.15mm, toprule=.15mm, bottomtitle=1mm, left=2mm, rightrule=.15mm, coltitle=black, titlerule=0mm, toptitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, colback=white]

@@

\begin{equation}\phantomsection\label{eq-squintability}{
\varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-logistic}{
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-parametric}{
f(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-squintability-parametric}{
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}
}\end{equation}

@@

\subsection{Performance of JSO in pipe-finding
problems}\label{sec-app-1}

@@

\subsection{Factors affecting JSO success rate: optimisation properties
and jellyfish hyper-parameters}\label{sec-app-2}

@@

\end{figure}%

\begin{table}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The column ``smooth'' and
``squint'' represent the smoothness and squintability measures.}

\centering{

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 3.053\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 2.960\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 2.948\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 3.117\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 1.257\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.772\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 1.322\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 1.369\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 2.947\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 2.756\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 3.118\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 2.957\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

@@

\begin{longtable}[t]{lrrrrrrl}

\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed
for modelling.}

\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

\end{longtable}

\endgroup{}

@@

\begin{longtable}[t]{|>{}lrrr>{}r|}

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on optimisation properties and jellyfish
hyper-parameters from the generalised linear model with a binomial
family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum number of tries are
positively associated with JSO success rate while data dimension and
being flagged as long runtime are negatively associated with the success
rate. The variable squintability and dimension are significant,
suggesting their importance relative to jellyfish hyper-parameters in
the optimisation success.}

\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 2.060 & 0.684 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule

\end{longtable}

\endgroup{}

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JSO success rate, JSO hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the JSO hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable \texttt{long\_time} to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, because upon visual
inspection of the final projection, none of the 50 simulations
identified the sine-wave shape. A generalised linear model with a
binomial family and a logit link function is used to fit the data and
Table~\ref{tbl-mod-output} presents the model outputs. The model
suggests that the JSO success rate is positively associated with the two
JSO hyper-parameters, as well as with the optimisation properties --
smoothness and squintability. Specifically, using 10 more jellyfish and
10 more tries will increase the odd ratio of success by 24.11\% and
11.93\%, respectively. However, being flagged with long runtime and an
increase of data dimension will reduce the success rate by 41.72\% and
53.36\%, respectively. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to JSO hyper-parameters in the optimisation success. Users can increase
the number of jellyfish and the number of tries to improve the success
rate, until JSO exceeds a 30-second runtime.

\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}
